Project Description -- Erosion, Transport, Anomalous Diffusion, and Collapse in Granular and Porous Media

### 2019-11-15 docs/proposals/NSF/ProjDesc/ProjDesc.tex
worked on some rearranging and added todos

An immediate goal of the proposed work is extend the range of physical effects captured in these simulations. A particular challenge of a BIE formulation is developing quadrature for nearly-singular integrands that arise when bodies are close. To address this problem, we have very recently extended a Barycentric quadrature~\cite{bar2014, bar-wu-vee2015} and have successfully simulated erosion with grains that are initially separated by 1\% of an arclength spacing. This quadrature approach has the advantage that it is a non-intrusive modification of the trapezoid rule, allowing us to achieve an error bound that independent of the grain configuration. Near-term plans are to include flow transport and gravitational sedimentation of the eroding grains. For this extension, contact forces between bodies must be included to eliminate unphysical contact that is inevitable due to discretization errors.

Though conceptually straightforward, implementing contact forces in multibody fluid-structure simulations is far from simple in practice, with several competing possible strategies that are a topic of current research.  One approach is to use introduce a potential with short-range repulsion~\cite{}, but such methods often introduce stiffness. Alternatively, a space-time interference volume can guarantee a minimum separation between grains.  In a fluid context, this method was first introduced by Lu et al.~\cite{lu-rah-zor2017}, and extended to allow for much smaller separation distances of rigid bodies by the PI's previous graduate student~\cite{bys-sha-qua2019}. One Ph.D.~student, who will be jointly supervised by both PIs, will simulate sedimentation of eroding bodies by developing algorithms that resolve quadrature error and contact.

Next, we plan to extend the simulation framework to handle chemical dissolution of solid bodies via physical laws \eqref{Ceq}--\eqref{DissVn}. For this goal, much of the computational infrastructure, such as the boundary-integral solver for the Stokes equations, is already in place, but the coupling to the advection-diffusion equations~\eqref{Ceq} and interface law~\eqref{DissVn} would need to be implemented. Once these pieces are in place, simulations of simultaneous erosion and dissolution, while also including transport and sedimentation, would finally become possible. Such a framework could accurately simulate the geophysically relevant scenario of karst conduits with embedded granular media. The gradual changes in the medium and karst boundaries due to erosion and dissolution could create conditions favorable for mechanical failure or collapse of the structure, for example sinkholes.
\todo[inline]{If we are going to propose the Laplace transform idea
to solve advection-diffusion, this might be the place to put it}

@@

The rate of dispersion is known to be directly related to the pore size distribution~\cite{dea-qua-bir-jua2018}. Therefore, there is a need to understand how erosion affects the pore size distribution. In preliminary, we define pore sizes between grains that share an edge of a Delanuay triangulation with vertices at the grain centers (Figure~\ref{fig:Eroding100gap_mean_var}(a)). In Figures~\ref{fig:Eroding100gap_mean_var}(a) and~\ref{fig:Eroding100gap_mean_var}(b), we plot the mean and variance of the pore sizes as a function of porosity. To better understand how erosion affects the anomalous dispersion, we propose to investigate how the pores grow with erosion. We will start with the notion of a {\em porelet} which is an individual Hagen-Poiseuille flow in each of the individual pores~\cite{dea-qua-bib-jua2018}. By computing the strength of the individual Hagen-Poiseuille flows, the shear rate can be computed, and thus the rate of erosion can be computed.

@@

We propose to extend our preliminary results by considering
concentrations with finite Peclet numbers. Numerical methods to solve
the advection-diffusion equation~\eqref{eqn:advectionDiffusion} in a
complex geometry poses several challenges that we will address.  We will
start by splitting the advection and diffusion equations with a Strang
splitting method~\cite{str1968}.  Numerical solutions of the advection
equation will be formed with a semi-Lagrangian method~\cite{rob1981} so
that it can be solved on an Eulerian grid without a restrictive CFL
condition. PI Quaife has experience with using semi-Lagrangian
methods for another type of complex viscous flow~\cite{kab-qua-bir2017}.
The diffusion equation is more challenging to solve in a complex
geometry with an integral equation formulation.  One approach is to use
the space-time heat kernel~\cite{}, but this requires storing an entire
time history of the solution.  Alternatively, the heat equation can be
discretized in time and the resulting Yukawa equation is solved at each
time step~\cite{kro-qua2010}. A particular challenge of this method is
extending the forcing function from the complex geometry and into a
regular geometry, such as a square, so that a high-order volume integral
or Fourier method can be applied.  Recently, groups have used radial
basis functions to form a sufficiently smooth
extension~\cite{fry-kro-tor2019, fry-leh-tor2019}.

@@

@@

	More specifically, we plan to run large batches of the high-fidelity boundary-integral-equation-based numerical simulations of erosion and then leverage rapid advances in the field of machine learning and deep-neural networks to develop reduced models that parametrize the effects of erosion on the evolution of macroscopic medium properties. The output of the high-fidelity simulations can be transformed into essential an image format, enabling us to use cutting-edge advancements on deep neural-networks for image analysis \cite{pelt2018mixed}. Some important advantages of referenced work are the use of dilated convolutions in a deeply connected network, which first mixes scales and second reduces the number of super-parameters present and thus largely averts the risk of overfitting data. Both PIs will supervise a graduate student who will design and implement the deep-neural network for learning the essence of the high-fidelity simulations. Once the essential features are learned, they will be incorporated into reduced-order models that utilize the multiphase framework to enable more efficient and larger scale coarse-grained simulations of dynamic erosion.

>>>>

An immediate goal of the proposed work is extend the range of physical effects captured in these simulations. A particular challenge of a BIE formulation is developing quadrature for nearly-singular integrands that arise when bodies are close. To address this problem, we have very recently extended a Barycentric quadrature~\cite{bar2014, bar-wu-vee2015} and have successfully simulated erosion with grains that are initially separated by 1\% of an arclength spacing~\cite{chi-moo-qua2019}. This quadrature approach has the advantage that it is a non-intrusive modification of the trapezoid rule, allowing us to achieve an error bound that independent of the grain configuration. Near-term plans are to include flow transport and gravitational sedimentation of the eroding grains. For this extension, contact forces between bodies must be included to eliminate unphysical contact that is inevitable due to discretization errors.

Though conceptually straightforward, implementing contact forces in multibody fluid-structure simulations is far from simple in practice, with several competing possible strategies that are a topic of current research.  One approach is to use introduce a potential with short-range repulsion, but such methods often introduce stiffness. Alternatively, a space-time interference volume can guarantee a minimum separation between grains.  In a fluid context, this method was first introduced by Lu et al.~\cite{lu-rah-zor2017}, and extended to allow for much smaller separation distances of rigid bodies by the PI's previous graduate student~\cite{bys-sha-qua2019}. One Ph.D.~student, who will be jointly supervised by both PIs, will simulate sedimentation of eroding bodies by developing algorithms that resolve quadrature error and contact.

Next, we plan to extend the simulation framework to handle chemical dissolution of solid bodies via physical laws~\eqref{Ceq}--\eqref{DissVn}. For this goal, much of the computational infrastructure, such as the boundary-integral solver for the Stokes equations, is already in place, but the coupling to the advection-diffusion equations~\eqref{Ceq} and interface law~\eqref{DissVn} need to be implemented.  We proposed to solve the advection diffusion equation~\eqref{Ceq} by first using Strang splitting~\cite{str1968} to decompose the PDE into linear advection and diffusion.  The advection equation will be solved with a semi-Lagrangian method~\cite{rob1981} so that it can be solved on an Eulerian grid without a restrictive CFL condition. PI Quaife has experience with using semi-Lagrangian methods for another type of complex viscous flow~\cite{kab-qua-bir2017}.  The diffusion equation is more challenging to accurately solve in the complex porous geometries.  Two approaches include a memory-intensive space-time heat kernel~\cite{bar-eps-gre-jia-wan2019, jia-gre-wan2015, li-gre2009}, and Rothe's method coupled with a Yukawa equation solver~\cite{kro-qua2010, cau-cho-chr-sea2016}. A particular challenge of the latter method is extending functions from the complex geometry and into a regular geometry, such as a square, so that a high-order volume integral or Fourier method can be applied, and some groups have recently developed new techniques to form such an extension~\cite{fry-kro-tor2019, fry-leh-tor2019}. Alternatively, we propose to use a Laplace Transform to solve the diffusion equation
\begin{alignat}{3}
  \pd{c}{t} &= \frac{1}{\Pe} \Delta c, \quad &&\xx \in \Omega, \: t>0, .
\end{alignat}
Applying the Laplace transform, $C(\xx,s) = \mathcal{L}[c](\xx,t)$ satisfies
\begin{align}
  sC(\xx,s) - \frac{1}{\Pe}\Delta C(\xx,s) = c_0(\xx), 
    \quad \xx \in \Omega,
\end{align}
where $c_0(\xx)$ is the initial concentration. This PDE will be solved with high-order accuracy and efficiently by using fast-multipole-accelerated integral equation methods of the Yukawa equation~\cite{kro-qua2011}, and the concentration will be recovered with an inverse Laplace transform~\cite{jos-war2006}.  
Once a high-order advection-diffusion solver is in place, by combining it with our integral equation formulation for the Stokes equation and the shear stress, simulations of simultaneous erosion and dissolution, while also including transport and sedimentation, would finally become possible. Such a framework could accurately simulate the geophysically relevant scenario of karst conduits with embedded granular media. The gradual changes in the medium and karst boundaries due to erosion and dissolution could create conditions favorable for mechanical failure or collapse of the structure, for example sinkholes.

@@

The rate of dispersion is known to be directly related to the pore size distribution~\cite{dea-qua-bir-jua2018}. Therefore, there is a need to understand how erosion affects the pore size distribution. In preliminary, we define pore sizes between grains that share an edge of a Delanuay triangulation with vertices at the grain centers (Figure~\ref{fig:Eroding100gap_mean_var}(a)). In Figures~\ref{fig:Eroding100gap_mean_var}(a) and~\ref{fig:Eroding100gap_mean_var}(b), we plot the mean and variance of the pore sizes as a function of porosity. To better understand how erosion affects the anomalous dispersion, we propose to investigate how the pores grow with erosion. We will start with the notion of a {\em porelet} which is an individual Hagen-Poiseuille flow in each of the individual pores~\cite{dea-qua-bir-jua2018}. By computing the strength of the individual Hagen-Poiseuille flows, the shear rate can be computed, and thus the rate of erosion can be computed.

@@

\todo[inline]{Advection diffusion done with the same method used to do
dissolution. A lot of the next few paragraphs will be simplified}
We propose to extend our preliminary results by considering
concentrations with finite Peclet numbers. Numerical methods to solve
the advection-diffusion equation~\eqref{eqn:advectionDiffusion} in a
complex geometry poses several challenges that we will address.  We will
start by splitting the advection and diffusion equations with a Strang
splitting method~\cite{str1968}.  Numerical solutions of the advection
equation will be formed with a semi-Lagrangian method~\cite{rob1981} so
that it can be solved on an Eulerian grid without a restrictive CFL
condition. PI Quaife has experience with using semi-Lagrangian
methods for another type of complex viscous flow~\cite{kab-qua-bir2017}.
The diffusion equation is more challenging to solve in a complex
geometry with an integral equation formulation.  One approach is to use
the space-time heat kernel~\cite{}, but this requires storing an entire
time history of the solution.  Alternatively, the heat equation can be
discretized in time and the resulting Yukawa equation is solved at each
time step~\cite{kro-qua2010}. A particular challenge of this method is
extending the forcing function from the complex geometry and into a
regular geometry, such as a square, so that a high-order volume integral
or Fourier method can be applied.  Recently, groups have used radial
basis functions to form a sufficiently smooth
extension~\cite{fry-kro-tor2019, fry-leh-tor2019}.

@@

\todo[inline]{Talk about how to develop models for the pore sizes as a
function of time. See results from S-H}

@@

	More specifically, we plan to run large batches of the high-fidelity boundary-integral-equation-based numerical simulations of erosion and then leverage rapid advances in the field of machine learning and deep-neural networks to develop reduced models that parametrize the effects of erosion on the evolution of macroscopic medium properties. The output of the high-fidelity simulations can be transformed into essential an image format, enabling us to use cutting-edge advancements on deep neural-networks for image analysis \cite{pelt2018mixed}. Some important advantages of referenced work are the use of dilated convolutions in a deeply connected network, which first mixes scales and second reduces the number of super-parameters present and thus largely averts the risk of overfitting data. Both PIs will supervise a graduate student who will design and implement the deep-neural network for learning the essence of the high-fidelity simulations. Once the essential features are learned, they will be incorporated into reduced-order models that utilize the multiphase framework to enable more efficient and larger scale coarse-grained simulations of dynamic erosion.
        \todo[inline]{Nathan Kutz has a 2017 JFM 4 page paper that talks
        about neural networks/deep learning in fluid dynamics}
